Solving the Nonstationary Richards Equation\\ With Adaptive $hp$-FEM

### 2011-04-29 paper.tex
fixed references + highlights

\address{${}^1$Department of Mathematics and Statistics, University of Nevada, Reno, USA}
\address{${}^2$Institute of Themomechanics, Prague, Czech Republic}
\address{${}^3$Department of Water Resources and Environmental Modeling, Czech University of Life Sciences, Prague, Czech Republic}
%\address{${}^4$Institute of Physics, Academy of Sciences of the Czech Republic, Czech Republic}

@@

The domain consists of two 50 cm thick surface layers and a low-permeability
clay block. The unsaturated material properties were obtained from~\cite{unsat-prop} 
and the saturated material properties (storativity) \cite{pech}. 
The first layer is a silt loam, and the second is a silty clay 
loam. Layer three is sandy clay loam and layer four silty clay again. 
For details on the material parameters see Fig.~\ref{fig:kees-uloha} 
and Table~\ref{Material-properties}. For the corresponding constitutive 
relations we use the van Genuchten's law~(\ref{vangenuchten-eq}) and Mualem's 
law~(\ref{mualem-int}).

\begin{table*}[!ht]
\begin{center}
\label{Material-properties}
\caption{Material properties.}
\begin{small}
% double K_S_1 = 0.108;
% double K_S_3 = 0.0048;
% double K_S_2 = 0.0168;
% //double K_S_4 = 41.143; 
% double K_S_4 = 1.061;
% 
% 
\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
type & $K_S$ [m.d${}^{-1}$] & $\theta_s$ [-] & $\theta_r$ [-] & $\alpha$ [m${}^{-1}$] & $n$ [-] & $m$ [-] & S [m${}^{-1}$] \\ \hline \hline
1 & 0.108 & 0.4670 & 0.1201 &0.15 &1.8 & 0.8 & 0.01330 \\\hline
2 & 0.0168 & 0.3205 & 0.1012 & 0.23 & 1.85 & 0.589 & 0.01262  \\\hline
3 & 0.0048 & 0.15 & 0.064 & 0.185 & 1.56 & 0.756 & 0.011955 \\\hline
4 & 1.061 & 0.3250 & 0.08590 & 1.85 & 2.0 & 0.055 & 0.012577  \\\hline
\end{tabular}
\end{small}
\end{center} 
\end{table*}

@@

For the time-discretization we have chosen the second-order accurate 
and unconditionally stable Crank-Nicolson method with a 
time step of 1.5 minutes. A few snapshots of the solution and meshes 
are shown in Figs. \ref{fig:see-sol-0} and \ref{fig:see-mesh-4}.

@@

\subsubsection{Period 2}
This is the most demanding part of the computation since the capillary barrier collapses 
and the gravel layer is quickly filled with the infiltrated water. The flow is dominantly 
vertical (see Figs.~\ref{fig:solution2.1} and~\ref{fig:solution2.2}) and thus the convective 
term $\frac{\partial K(h)}{\partial z}$ attains large values. This fact, together with an 
extremely nonlinear nature of the unsaturated 
hydraulic functions of the gravel close to full saturation,
a very fine mesh develops at the wetting front.

@@

AGROS (http://hpfem.org/agros2d) is a multiplatform C++ application for the solution 
of single and multiphysics PDE problems based on the HERMES library. AGROS provides 
an advanced graphical user interface (GUI), interactive geometry modeling and CAD import, 
automatic meshing, and a number of advanced visualization and postprocessing features. 
The code is being developed by an engineering research group at the University of West 
Bohemia in Pilsen, Czech Republic. AGROS is distributed under the GNU General Public 
License. Currently, it supports the finite element analysis of electrostatic fields,
electric current fields, magnetic fields (steady state, harmonic and transient analysis),
heat transfer (steady state and transient analysis), basic fluid mechanics and linear 
elasticity (in development). A Richards module for variably saturated Darcian %odstranil jsem groundwater
flow is currently in development.

>>>>

\address{${}^1$Department of Mathematics and Statistics, University of Nevada, Reno, USA}
\address{${}^2$Institute of Themomechanics, Prague, Czech Republic}
\address{${}^3$Department of Water Resources and Environmental Modeling, Czech University of Life Sciences Prague, Czech Republic}
%\address{${}^4$Institute of Physics, Academy of Sciences of the Czech Republic, Czech Republic}

@@

The domain consists of two 50 cm thick surface layers and a low-permeability
clay block. The unsaturated material properties were obtained from~\cite{unsat-prop} 
and the saturated material properties (storativity) \cite{pech}. 
The first layer is a silt loam, and the second is a silty clay 
loam. Layer three is sandy clay loam and layer four silty clay again. 
For details on the material parameters see Fig.~\ref{fig:kees-uloha} 
and Table~2. For the corresponding constitutive 
relations we use the van Genuchten's law~(\ref{vangenuchten-eq}) and Mualem's 
law~(\ref{mualem-int}).

\begin{table*}[!ht]
\begin{center}
\caption{Material properties.}
\begin{small}
% double K_S_1 = 0.108;
% double K_S_3 = 0.0048;
% double K_S_2 = 0.0168;
% //double K_S_4 = 41.143; 
% double K_S_4 = 1.061;
% 
% 
\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
type & $K_S$ [m.d${}^{-1}$] & $\theta_s$ [-] & $\theta_r$ [-] & $\alpha$ [m${}^{-1}$] & $n$ [-] & $m$ [-] & S [m${}^{-1}$] \\ \hline \hline
1 & 0.108 & 0.4670 & 0.1201 &0.15 &1.8 & 0.8 & 0.01330 \\\hline
2 & 0.0168 & 0.3205 & 0.1012 & 0.23 & 1.85 & 0.589 & 0.01262  \\\hline
3 & 0.0048 & 0.15 & 0.064 & 0.185 & 1.56 & 0.756 & 0.011955 \\\hline
4 & 1.061 & 0.3250 & 0.08590 & 1.85 & 2.0 & 0.055 & 0.012577  \\\hline
\end{tabular}
\end{small}
\end{center} 
\label{table:2}
\end{table*}

@@

For the time-discretization we have chosen the second-order accurate 
and unconditionally stable Crank-Nicolson method with a 
time step of 1.5 minutes. A few snapshots of the solution and meshes 
are shown in Figs. \ref{fig:see-sol-0}, \ref{fig:see-mesh-0}, \ref{fig:see-sol-1}, \ref{fig:see-mesh-1}, \ref{fig:see-sol-4}, and \ref{fig:see-mesh-4}.

@@

\subsubsection{Period 2}
This is the most demanding part of the computation since the capillary barrier collapses 
and the gravel layer is quickly filled with the infiltrated water. The flow is dominantly 
vertical (see Figs.~\ref{fig:solution2.1}, \ref{fig:solution2.2}, and~\ref{fig:solution2-mesh}) and thus the convective 
term $\frac{\partial K(h)}{\partial z}$ attains large values. This fact, together with an 
extremely nonlinear nature of the unsaturated 
hydraulic functions of the gravel close to full saturation,
a very fine mesh develops at the wetting front.

@@

AGROS (http://hpfem.org/agros2d) is a multiplatform C++ application for the solution 
of single and multiphysics PDE problems based on the HERMES library. AGROS provides 
an advanced graphical user interface (GUI), interactive geometry modeling and CAD import, 
automatic meshing, and a number of advanced visualization and postprocessing features (see Fig.~\ref{fig:agros}). 
The code is being developed by an engineering research group at the University of West 
Bohemia in Pilsen, Czech Republic. AGROS is distributed under the GNU General Public 
License. Currently, it supports the finite element analysis of electrostatic fields,
electric current fields, magnetic fields (steady state, harmonic and transient analysis),
heat transfer (steady state and transient analysis), basic fluid mechanics and linear 
elasticity (in development). A Richards module for variably saturated Darcian %odstranil jsem groundwater
flow is currently in development.
